---
title: "PS270 Final Project"
author: "Sam Schallert"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to sports and political voting based on party/ideals. Both these topics intrest me and I feel there is a lot of data about each of them. 

## Milestone 2: Research Proposal
 

This project utilizes the "NFL Scores and Betting Data" dataset to investigate the relationship between market expectations and actual game outcomes. My research question asks: To what extent does the closing point spread set by oddsmakers predict the actual margin of victory in NFL games? I hypothesize that there is a strong, positive linear relationship between the spread and the final score margin. This hypothesis is rooted in the Efficient Market Hypothesis, which suggests that in a competitive environment with high financial stakes, betting lines will systematically incorporate all publicly available information regarding team strength, injuries, and conditions. My explanatory variable is the spread_favorite, which measures the number of points the favored team is expected to win by (ranging from -1 to -26.5 in this data). My outcome variable is the actual_margin, calculated as the home team's score minus the away team's score. A statistically significant, positive regression coefficient near 1.0 would provide support for my hypothesis, indicating that the market is a reliable predictor of reality. Conversely, a coefficient near zero, or a high degree of variance in which the actual scores frequently deviate widely from the spread, would disprove the hypothesis and suggest that NFL outcomes are too idiosyncratic for systematic market prediction.


```{r}
# 1. Load the necessary library
library(tidyverse)

# 2. Load the data from your 'data' folder
nfl_data <- read_csv("data/spreadspoke_scores.csv")

# 3. Meaningful Data Cleaning
# We filter out games that haven't happened yet (NA scores) 
# and create our outcome variable: actual_margin
nfl_clean <- nfl_data %>%
  filter(!is.na(score_home) & !is.na(score_away)) %>%
  mutate(actual_margin = score_home - score_away) %>%
  # Selecting just a few columns for the head() so it looks clean on your site
  select(schedule_date, team_home, team_away, score_home, score_away, spread_favorite, actual_margin)

# 4. Print the head of the dataset
head(nfl_clean)
```

## Milestone 3:Data Cleanup

```{r nfl-visualization, echo=FALSE, message=FALSE, warning=FALSE}
# 1. Data Cleaning (Filtering for complete games and claulating margin)
library(tidyverse)
nfl_vis_data <- nfl_data %>%
  filter(!is.na(score_home) & !is.na(spread_favorite)) %>%
  mutate(actual_margin = score_home - score_away)

# 2. Create a Polished Visualization
ggplot(nfl_vis_data, aes(x = spread_favorite, y = actual_margin)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Market Accuracy: NFL Point Spreads vs Actual Margins",
    subtitle = "Data from 1966 to present",
    x = "Betting Spread (Points for Favorite)",
    y = "Actual Source Margin (Home - Away)", 
    caption = "Source: NFL Scores and Betting Data(Kaggle)"
  ) +
  theme_minimal() +
  theme(plot.tittle = element_text(face = "bold", hjust = 0.5))
```

To prepare this data for analysis, I performed several cleaning steps to ensure accuracy. First, I filtered the dataset to remove any games with missing scores or missing betting spreads, ensuring the regression model would only utilize complete observations. I then used a mutation to calculate the actual_margin of each game (Home Score minus Away Score), which serves as my dependent variable. The visualization below displays a scatter plot of the spread_favorite against the actual_margin. I have included a linear regression line (red) to illustrate the systematic relationship between these two variables. This "polished" visualization allows us to see the density of the data: while there is a clear positive correlation (supporting the idea that oddsmakers are generally accurate), the high degree of "noise" or "spread" around the line suggests that individual NFL games frequently deviate from market expectations due to idiosyncratic factors like turnovers or weather.



